# Hampton Roads {#part-2-1}

```{r}

# Describe the Hampton Roads region. 

# What has the housing market been like in recent years? Use news articles and available data to summarize trends.

# Use data from the VZA to summarize the number of districts involved, number of zoning ordinance pages read, and other interesting metrics.

```

The [Hampton Roads Planning District Commission](https://www.hrpdcva.gov/) is comprised of 17 local governments, including ten independent cities, six counties, and the Town of Smithfield. This analysis also includes the Towns of Windsor and Surry. The other towns in the region, Boykins, Branchville, Capron, Claremont, Courtland, Dendron, Ivor, and Newsoms, were not part of this analysis because their zoning map and/or zoning ordinance were unavailable to us.

```{r}
#| label: fig-hr-map
#| fig-cap: "Hampton Roads"

library(sf)
library(leaflet)
library(readr)

hr_region <- read_rds("data/hr/hr_region.rds")

leaflet(hr_region) |> 
  addTiles() |> 
  addPolygons()


```

## Housing Market



[Summarize some key facts about the Hampton Roads housing market and the region as a whole]

https://whro.org/news/local-news/36579-hampton-roads-housing-market-due-for-a-slowdown-says-odu

https://www.vpm.org/news/2022-11-07/housing-in-hampton-roads-is-less-affordable-than-northern-virginia-other-pricey

https://whro.org/news/local-news/33171-analyzing-hampton-roads-affordability-a-conversation-with-journalist-jeff-south


## Data Collection

Zoning data collection and analysis for the Hampton Roads region took place from April through August 2023. At this time, Gloucester County and the City of Poquoson were going through updates of their zoning ordinances. The findings and presentation of the Hampton Roads portion of the Virginia Zoning Atlas are a point-in-time representation of zoning ordinances during this time.

It is the intent of the Virginia Zoning Atlas to track major changes to zoning ordinances and zoning maps with periodic updates.

## 