# Zoning in Hampton Roads {part-2-2}

```{r}
#| label: setup

library(tidyverse)
library(tidycensus)
library(ggplot2)
library(ggiraph)
library(sf)
library(leaflet)
library(FinCal)
library(hdatools)
library(scales)
library(forcats)
library(kableExtra)
library(ggtext)
# Summarize the key findings from the analysis below.

# Describe the housing market in relation to population growth and household trends in the region. 

# Is the population growing or shrinking? What is the typical household size?
# Are we building the housing to match the growth and typical household?

# Describe economic trends in the region.

# Is there job growth in the region? Are incomes and wages growing or declining?
# What can the typical renter afford to buy or rent?
# What can the top occupations afford to buy or rent?

# Describe zoning's impact in Hampton Roads.

# How much of zoned land in Hampton Roads is dedicated to single-family housing only (% of total area)? Total and by locality. Ranked.
# How much of zoned land in Hampton Roads is primarily residential, mixed with residential, and nonresidential (% of total area)? Total and by locality. Ranked.
# How much of zoned land in Hampton Roads allows for 2+ family by-right (% of total area)?
# Housing type by-right by locality
# Where are there minimum square footage requirements?
# How much of zoned land in Hampton Roads allows for ADUs by-right?
# What is the average minimum lot area for each treatment?


```
### Data Collection

Zoning data collection and analysis for the Hampton Roads region took place from April through August 2023. At this time, Gloucester County and the City of Poquoson were going through updates of their zoning ordinances. The findings and presentation of the Hampton Roads portion of the Virginia Zoning Atlas are a point-in-time representation of zoning ordinances during this time. Zoning and geospatial analysis contributions were made by Christopher Newport University students: Robert Bennett, Stephen Everard, Audrey Morrison, Orson Lange, and Layne Marshall. Eric Mai and Maria Dougherty were the project leads for the region.

It is the intent of the Virginia Zoning Atlas to track major changes to zoning ordinances and zoning maps with periodic updates.

## Key Findings

- Mixed with Residential zoning districts make up a small percentage of zoned land in most jurisdictions.
- 
- 


## Zoning's impact in Hampton Roads

With increasing housing affordability issues across the region, a diverse and growing supply of housing is needed to meet the needs of a diverse population. Localities have significant control over their supply through their zoning ordinances. 

The following findings are based on the Zoning Atlas Methodology and takes into account the impact of federally-owned lands and wetland areas, which are both prevalent throughout the region. Like our colleagues in New Hampshire, we have sought to exclude areas of land that are not unbuildable. We have excluded all federal land from calculations, as well as area of wetlands and/or water that are 10 contiguous area or more.

::: {.callout-note}

*Federal lands* were identified using ESRI Landscape's USA Federal Lands feature layer, which accounted for lands managed by six federal agencies, including: 

- Bureau of Land Management
- Department of Defense
- National Park Service
- US Fish and Wildlife Service
- US Forest Service

*Wetland/water areas* were identified using the U.S. Fish & Wildlife Service's Seamless Wetlands Data. The wetlands polygon was used and allow polygons with an area of 10 acres or more were utilized as a mask layer to clip from our Hampton Roads zoning district layer file.

:::

Overlay districts, which are districts that may change the regulations of a base zoning district, were not a part of this analysis. Few overlay districts in the Hampton Roads region affect the uses of the base zoning district. 

```{r}
#| label: tbl-hr-zoning-summary
#| tbl-cap: "Summary of zoning ordinances in Hampton Roads"

# Read in csv of zoning summary for Hampton Roads.
hr_zoning_summary <- read_csv("data/hr/hr_zoning_summary.csv")


hr_zoning_summary |> 
  kable(col.names = c("Jurisdiction", "County or equivalent", "Government Type", "Date of Zoning Ordinance", "Number of Pages", "Number of base and overlay districts"), align = "lllccc") |> 
  kable_styling(bootstrap_options = c("striped", "condensed"))


```

Zoning ordinances see amendments made over the years that eliminate, add, or refine language to better meet the needs of the community. But comprehensive rewrites following a comprehensive planning process can help ensure that zoning regulations align with future land use goals and changing demographics. From the table above, you can see that several localities have updated their zoning ordinances in recent years, while others have not had comprehensive update in decades.

## Type of Zoning District

```{r}
#| label: fig-housing-type
#| fig-cap: "Percent of land zoned by residential category"

# Create a bar char that shows the percentage of zoned in each locality by type (i.e. primarily residential, mixed with residential, nonresidential)

hr_type <- read_rds("data/hr/hr_vza_nogeo.rds") |> 
  select(jurisdiction, overlay, type, area) |> 
  filter(overlay == 0)

hr_type_summary <- hr_type |> 
  group_by(jurisdiction, type) |> 
  summarise(area = sum(area)) |> 
  group_by(jurisdiction) |> 
  mutate(percent = area/sum(area))

hr_summary <- hr_type |> 
  group_by(type) |> 
  summarise(area = sum(area)) |> 
  mutate(percent = area/sum(area))


ggplot(hr_type_summary,
       aes(x = fct_reorder2(jurisdiction, type, percent),
           y = percent,
           fill = factor(type, levels = c("Nonresidential", "Mixed with Residential", "Primarily Residential")))) +
  geom_col(position = "stack") +
  coord_flip() +
  theme_hfv() +
  scale_fill_hfv() +
  theme(legend.position = "right") +
  labs(title = "Percent of land by type of zoning district",
       caption = "**Source:** HousingForward Virginia, Virginia Zoning Atlas.") +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.01)))


```
Roughly 90 percent of developable land is **Primarily Residential**. This includes prime agricultural land in the southern parts of Hampton Roads, where the preservation of open space and natural resources is a vital part of the region. Only three percent of the region's developable land is classified as **Mixed with Residential**. Locally, Williamsburg (22%), York County (14%), and Newport News (13%) have the largest proportions of Mixed with Residential zoning types, while Surry County (0%), Southampton County (0%), and Isle of Wight County (less than 1%) have the least.

Walkable, mixed use communities are [increasingly desirable among all age groups](https://www.strongtowns.org/journal/2023/7/25/realtors-buyers-will-pay-extra-for-walkable-neighborhoods). The ability to reach work, shopping, and leisure activities within walking distance supports healthy communities, but also supports diverse local business and job opportunities (think of areas like like Ghent in Norfolk or the ViBe District in Virginia Beach). 


### Single-Family Detached Housing

While single-family detached housing is a dream many aspire to, barriers to homeownership are making it difficult for many Hampton Roads residents to purchase a home.The previous section highlighted those challenges with current wages and high rental costs that are further limiting the ability to save for a home purchase. Furthermore, single-family detached only zoning can have major implications on equity and economic opportunity. Allowing for more diverse types of housing in single-family detached only zoning districts can help provide greater access to high opportunity areas for a diverse range of households - both in terms of race and ethnicity, and income.

```{r}
#| label: fig-sfd-map
#| fig-cap: "Single-family detached only zoning in Hampton Roads"

# Create a map that shows where single-family detached only housing districts are in the region.


vza_pal <- colorFactor(palette =c("#8B85CA", "#40C0C0", "#011E41"),
                       levels = c("Primarily Residential", "Mixed with Residential", "Nonresidential"))

hr_sfd_map <- read_rds("data/hr/hr_vza_geo.rds") |>
  filter(sfd == "t")

library(leafgl)
library(sf)

hr_sfd_map <- st_cast(hr_sfd_map, "POLYGON")

leaflet() |>
  addTiles() |>
  addGlPolygons(data = hr_sfd_map,
              color = ~vza_pal(type),
              weight = 1,
              fillOpacity = 0.75,
              popup = paste0("Jurisdiction: ", hr_sfd_map$jurisdiction,
                             "<br>",
                             "District Abbreviation: ", hr_sfd_map$abbrvname)) |>
  addLegend(labels = c("Primarily Residential", "Mixed with Residential", "Nonresidential"),
            colors = c("#8B85CA", "#40C0C0", "#011E41"),
            title = "Type of Zoning District")

# write_rds(zd, "data/hr/hr_sfd_map.rds")
# 
# hr_sfd_map <- read_rds("data/hr/hr_sfd_map.rds")
# 
# hr_sfd_map |> 
#   addPolygons()


```

In most Hampton Roads localities, the majority of developable land is dedicated to single-family detached only. Many of these localities are largely rural localities, where single-family detached housing is more appropriate in rural setting. But even in more urban and suburban areas like Virginia Beach, Hampton, Norfolk, and Newport News, a large proportion of those localities is dedicated to single-family detached only.

```{r}
#| label: fig-sfd-bar
#| fig-cap: "Percent of developable land dedicated to single-family detached only"

# Create a bar char that shows the percentage of zoned in each locality that is dedicated to single-family detached only housing.

hr_sfd <- read_rds("data/hr/hr_vza_nogeo.rds") |> # Read in the data.
  select(jurisdiction, overlay, sfd, area) |> # Select necessary data fields.
  filter(overlay == 0) |> # Remove overlay districts.
  group_by(jurisdiction, sfd) |> # Sum the total acres dedicated to single-family detached in each district.
  summarise(area = sum(area)) |> 
  group_by(jurisdiction) |> 
  mutate(percent = area/sum(area)) |> # Calculate what percentage of total developable area is SFD.
  filter(sfd == "t") 



ggplot(hr_sfd,
       aes(x = reorder(jurisdiction, percent),
           y = percent,
           fill = percent)) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge") +
  geom_text(aes(label = percent(percent, 1), color = percent),
    hjust = -0.2) +
  theme_hfv() +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  scale_color_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  labs(title = "Percent of developable land dedicated to single-family detached only",
       caption = "**Source:** HousingForward Virginia, Virginia Zoning Atlas.") +
  theme(axis.text.x = element_blank())

```

```{r}
#| label: fig-bipoc-sfd
#| fig-cap: "Percentage of SFD-only land compared to BIPOC population"
#| eval: false


# Read in data frame 
# hr_bipoc <- read_rds("data/hr/hr_bipoc.rds") |> 
#   st_drop_geometry() |> 
#   arrange(name_long) |> 
#   mutate(bipoc = percent) |> 
#   select(fips, name_long, bipoc)
# 
# hr_sfd <- read_rds("data/hr/hr_vza_nogeo.rds") |> # Read in the data.
#   select(jurisdiction, overlay, sfd, area) |> # Select necessary data fields.
#   filter(overlay == 0) |> # Remove overlay districts.
#   mutate(jurisdiction = case_when(
#     jurisdiction == "Smithfield" ~ "Isle of Wight",
#     jurisdiction == "Windsor" ~ "Isle of Wight", 
#     jurisdiction == "Surry (town)" ~ "Surry",
#     TRUE ~ jurisdiction
#   )) |> 
#   group_by(jurisdiction, sfd) |> # Sum the total acres dedicated to single-family detached in each district.
#   summarise(area = sum(area)) |> 
#   group_by(jurisdiction) |> 
#   mutate(percent = area/sum(area)) |> # Calculate what percentage of total developable area is SFD.
#   filter(sfd == "t") |> 
#   ungroup() |> 
#   add_row(jurisdiction = "York", sfd = "f", area = 0, percent = 0)
# 
# hr_sfd_arrange <- hr_sfd |> 
#   arrange(jurisdiction)
# 
# sfd_bipoc <- cbind(hr_bipoc, hr_sfd_arrange) |> 
#   mutate(sfd = percent) |> 
#   select(jurisdiction, bipoc, sfd)
# 
# # |> 
# #   pivot_longer(4:5,
# #                names_to = "metric",
# #                values_to = "pct") |> 
# #   mutate(metric = case_when(
# #     metric == "bipoc" ~ "BIPOC population (% of total)",
# #     metric == "sfd" ~ "Single-family detached only (% of total developable)"
# #   ))
#   
# 
# ggplot(sfd_bipoc,
#        aes(x = bipoc,
#            y = sfd)) +
#   geom_point() +
#   geom_smooth(method=lm)
# 
# +
#   coord_flip() +
#   theme_hfv() +
#   scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.05))) +
#   theme(legend.position = "top") + 
#   scale_fill_hfv() +
#   labs(title = "Percentage of SFD-only land compared to BIPOC population")



```
### By-Right Housing

When housing development requires a public hearing process, it can often come up against opposition from existing community members who have a preconceived notions about the impacts of increased density or new neighbors. The entitlement process, whether through a special use permit (or conditional use permit), variance, or rezoning can increase development costs and lead to project uncertainty. At a time when new housing is needed across the region, allowing for housing by-right (without a public hearing process) can streamline development and ensure that the housing supply is aligned with the demand.

```{r}
#| label: tbl-hr-byright
#| tbl-cap: "Percent of developable land that welcomes housing type by-right"

# Create a table that shows the percentage of land that allows for each type of treatment by locality
# https://frontierinstitute.org/reports/the-montana-zoning-atlas-2-0/ use this as an example 

hr_byright_tbl <- read_rds("data/hr/hr_vza_nogeo.rds") |> 
  select(jurisdiction, abbrvname, overlay, 6:9, area) |> 
  filter(overlay == 0) |> 
  group_by(jurisdiction) |> 
  mutate(total_area = sum(area)) |> 
  mutate(family1_pct = case_when(
    family1_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0
  ))|> 
  mutate(family2_pct = case_when(
    family2_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0 
  )) |> 
  mutate(family3_pct = case_when(
    family3_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0 
  )) |> 
  mutate(family4_pct = case_when(
    family4_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0 
  )) |> 
  group_by(jurisdiction) |> 
  summarise(family1_byright = percent(sum(family1_pct)),
            family2_byright = percent(sum(family2_pct)),
            family3_byright = percent(sum(family3_pct)),
            family4_byright = percent(sum(family4_pct))
            ) 


hr_byright_tbl |>
  kable(col.names = c("Jurisdiction", "Single-family", "2-family", "3-family", "4+-family"), 
        align = "lcccc") |> 
  kable_styling(bootstrap_options = c("striped", "condensed"))

```

Overwhelmingly, single-family is welcomed (without a public hearing) in nearly every locality by a wide margin, while 2+ family housing is only allowed by-right by less than seven percent in most localities. Areas like Newport News, Norfolk, and Williamsburg stand out as allowing much more than their neighbors. Portsmouth is notable for its allowance of duplexes and triplexes.
 
```{r}
#| label: fig-2plus
#| fig-cap: "Percent of developable land allowing for 2+ family housing by-right" 

# Create a bar graph that shows the total amount of zoned land that welcomes 2+ family housing by-right

hr_byright <- read_rds("data/hr/hr_vza_nogeo.rds") |> 
  select(jurisdiction, abbrvname, overlay, 6:9, area) |> # Select data fields.
  filter(overlay == 0) |> # Filter out overlay districts.
  group_by(jurisdiction) |> 
  mutate(total_area = sum(area)) |> # Calculate the total area of zoned land.
  filter(family2_treatment == "allowed" & family3_treatment == "allowed" & family4_treatment == "allowed") |> 
  group_by(jurisdiction) |> 
  mutate(pct_2plus = area/total_area) |> 
  group_by(jurisdiction) |> 
  summarise(pct_2plus = sum(pct_2plus)) |>
  add_row(jurisdiction = "Smithfield", pct_2plus = 0)
  

ggplot(hr_byright,
       aes(x = reorder(jurisdiction, pct_2plus),
           y = pct_2plus,
           fill = pct_2plus)) +
  geom_col() +
  geom_text(aes(label = percent(pct_2plus, 1), color = pct_2plus),
    hjust = -0.2) +
  coord_flip() +
  theme_hfv() +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  scale_color_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Percent of developable land allowing for 2+ family housing by-right")  +
  theme(axis.text.x = element_blank())




```

```{r}
#| label: fig-byr2-map
#| fig-cap: "Developable land zoned for 2+ family housing by-right"

# Create a map that shows where 2 + family housing is allowed by-right in the region.

pal <- colorFactor(palette =c("#8B85CA", "#40C0C0"),levels = c("R", "M"))

hr_byr2 <- read_rds("data/hr/hr_byr_map.rds")

hr_byr2 

```


### Minimum lot sizes

```{r}

# What is the average minimum lot size for each type of treatment? 
# Blank fields should be treated as NA and should not be included in the mean calculation.


hr_lots <- read_rds("data/hr/hr_vza_nogeo.rds")|> 
  select(jurisdiction, abbrvname, overlay, hr_custom_fields_min1, hr_custom_fields_min2,
         hr_custom_fields_min3, hr_custom_fields_min4) 


```


## Accessory Dwelling Units

```{r}
#| label: fig-hr-adus
#| fig-cap: "Percent of developable land that allows for ADUs by-right"

# Create a table that shows the percentage of land that allows for ADUs by-right
# https://frontierinstitute.org/reports/the-montana-zoning-atlas-2-0/ use this as an example 

hr_adu <- read_rds("data/hr/hr_vza_nogeo.rds") 

```


Many localities across Hampton Roads address ADUs in their zoning ordinances, while a few (Chesapeake and Newport News) make no mention of them at all. 
```{r}
#| label: fig-hr-adu-restrictions
#| fig-cap: "Summary of ADU regulations in Hampton Roads"

# Create a table for the following csv.
# https://frontierinstitute.org/reports/the-montana-zoning-atlas-2-0/ use this as an example 

hr_adu_summary <- read_csv("data/hr/hr_adu_summary.csv")

hr_adu_summary |> 
  kable() |> 
  kable_styling(bootstrap_options = c("striped", "condensed"))



```

## Access to Transit

Access to frequent and efficient public transportation can help entire regions reduce congestion, improve access to job opportunities, and make for more vibrant communities. By building diverse types of housing along transit lines, localities can better connect residents to opportunity and foster less of a reliance on personal vehicle travel.

[Transit oriented development (TOD)](https://www.transit.dot.gov/TOD) maximizes a mix of land uses within walking distance of public transit. Sustainable growth strategies, like TOD, are already utilized in Norfolk, as well as in Chesapeake, with their [Transportation Corridor Overlay District (TCOD)](https://www.cityofchesapeake.net/700/Transportation-Corridor-Overlay-District).

The following map shows zoned land that allows for 2+ family housing by-right in Hampton Roads with the location of Hampton Roads Transit bus stops. A quarter mile buffer, which is typically the minimum distance people will walk to a transit stop, are included.

```{r}
#| label: fig-transit-map
#| fig-cap: "Hampton Roads Transit bus stops and zoned land allowing for 2+ family housing by-right"

pal <- colorFactor(palette =c("#8B85CA", "#40C0C0"),levels = c("R", "M"))

hr_transit_map <- read_rds("data/hr/hr_transit_map.rds")

hr_transit_map



```
This map shows that many 2+ housing options are made available within bus stops, but there are still opportunities to increase transit access throughout the region by upzoning in strategic areas and/or expanding transit service.