# Zoning in Hampton Roads {part-2-2}

```{r}
#| label: setup

library(tidyverse)
library(tidycensus)
library(ggplot2)
library(ggiraph)
library(sf)
library(leaflet)
library(FinCal)
library(hdatools)
library(scales)
library(forcats)
library(kableExtra)
library(ggtext)
# Summarize the key findings from the analysis below.

# Describe the housing market in relation to population growth and household trends in the region. 

# Is the population growing or shrinking? What is the typical household size?
# Are we building the housing to match the growth and typical household?

# Describe economic trends in the region.

# Is there job growth in the region? Are incomes and wages growing or declining?
# What can the typical renter afford to buy or rent?
# What can the top occupations afford to buy or rent?

# Describe zoning's impact in Hampton Roads.

# How much of zoned land in Hampton Roads is dedicated to single-family housing only (% of total area)? Total and by locality. Ranked.
# How much of zoned land in Hampton Roads is primarily residential, mixed with residential, and nonresidential (% of total area)? Total and by locality. Ranked.
# How much of zoned land in Hampton Roads allows for 2+ family by-right (% of total area)?
# Housing type by-right by locality
# Where are there minimum square footage requirements?
# How much of zoned land in Hampton Roads allows for ADUs by-right?
# What is the average minimum lot area for each treatment?


```
### Data Collection

Zoning data collection and analysis for the Hampton Roads region took place from April through August 2023. At this time, Gloucester County and the City of Poquoson were going through updates of their zoning ordinances. The findings and presentation of the Hampton Roads portion of the Virginia Zoning Atlas are a point-in-time representation of zoning ordinances during this time. Zoning and geospatial analysis contributions were made by Robert Bennett, Stephen Everard, Audrey Morrison, Orson Lange, and Layne Marshall.

It is the intent of the Virginia Zoning Atlas to track major changes to zoning ordinances and zoning maps with periodic updates.

## Key Findings

- Mixed with Residential zoning districts make up a small percentage of zoned land in most jurisdictions.
- 
- 


## Zoning's impact in Hampton Roads

With increasing housing affordability issues across the region, a diverse and growing supply of housing is needed to meet the needs of a diverse population. Localities have significant control over their supply through their zoning ordinances. 

The following findings are based on the Zoning Atlas Methodology and takes into account the impact of federally-owned lands and wetland areas, which are both prevalent throughout the region. Like our colleagues in New Hampshire, we have sought to exclude areas of land that are not unbuildable. We have excluded all federal land from calculations, as well as area of wetlands and/or water that are 10 continguous area or more.

::: {.callout-note}

*Federal lands* were identified using ESRI Landscape's USA Federal Lands feature layer, which accounted for lands managed by six federal agencies, including: 

- Bureau of Land Management
- Department of Defense
- National Park Service
- US Fish and Wildlife Service
- US Forest Service

*Wetland/water areas* were identified using the U.S. Fish & Wildlife Service's Seamless Wetlands Data. The wetlands polygon was used and allow polygons with an area of 10 acres or more were utilized as a mask layer to clip from our Hampton Roads zoning district layer file.

:::

```{r}
#| label: fig-housing-type
#| fig-cap: "Percent of land zoned by residential category"

# Create a bar char that shows the percentage of zoned in each locality by type (i.e. primarily residential, mixed with residential, nonresidential)

hr_type <- read_rds("data/hr/hr_vza_nogeo.rds") |> 
  select(jurisdiction, overlay, type, area) |> 
  filter(overlay == 0)

hr_type_summary <- hr_type |> 
  group_by(jurisdiction, type) |> 
  summarise(area = sum(area)) |> 
  group_by(jurisdiction) |> 
  mutate(percent = area/sum(area))

ggplot(hr_type_summary,
       aes(x = fct_reorder2(jurisdiction, type, percent),
           y = percent,
           fill = factor(type, levels = c("Nonresidential", "Mixed with Residential", "Primarily Residential")))) +
  geom_col(position = "stack") +
  coord_flip() +
  theme_hfv() +
  scale_fill_hfv() +
  theme(legend.position = "right") +
  labs(title = "Percent of land by type of zoning district",
       caption = "*Source:** HousingForward Virginia, Virginia Zoning Atlas.") +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.01)))


```

### Single-Family Detached Housing


```{r}
#| label: fig-sfd-map
#| fig-cap: "Single-family detached only zoning in Hampton Roads"

# Create a map that shows where single-family detached only housing districts are in the region.


vza_pal <- colorFactor(palette =c("#8B85CA", "#40C0C0", "#011E41"),
                       levels = c("Primarily Residential", "Mixed with Residential", "Nonresidential"))

hr_sfd_map <- read_rds("data/hr/hr_vza_geo.rds") |>
  filter(sfd == "t")

library(leafgl)
library(sf)

hr_sfd_map <- st_cast(hr_sfd_map, "POLYGON")

leaflet() |>
  addTiles() |>
  addGlPolygons(data = hr_sfd_map,
              color = ~vza_pal(type),
              weight = 1,
              fillOpacity = 0.75,
              popup = paste0("Jurisdiction: ", hr_sfd_map$jurisdiction,
                             "<br>",
                             "District Abbreviation: ", hr_sfd_map$abbrvname)) |>
  addLegend(labels = c("Primarily Residential", "Mixed with Residential", "Nonresidential"),
            colors = c("#8B85CA", "#40C0C0", "#011E41"),
            title = "Type of Zoning District")

# write_rds(zd, "data/hr/hr_sfd_map.rds")
# 
# hr_sfd_map <- read_rds("data/hr/hr_sfd_map.rds")
# 
# hr_sfd_map |> 
#   addPolygons()


```

```{r}
#| label: fig-sfd-bar
#| fig-cap: "Percent of developable land dedicated to single-family detached only"

# Create a bar char that shows the percentage of zoned in each locality that is dedicated to single-family detached only housing.

hr_sfd <- read_rds("data/hr/hr_vza_nogeo.rds") |> # Read in the data.
  select(jurisdiction, overlay, sfd, area) |> # Select necessary data fields.
  filter(overlay == 0) |> # Remove overlay districts.
  group_by(jurisdiction, sfd) |> # Sum the total acres dedicated to single-family detached in each district.
  summarise(area = sum(area)) |> 
  group_by(jurisdiction) |> 
  mutate(percent = area/sum(area)) |> # Calculate what percentage of total developable area is SFD.
  filter(sfd == "t") 



ggplot(hr_sfd,
       aes(x = reorder(jurisdiction, percent),
           y = percent,
           fill = percent)) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge") +
  geom_text(aes(label = percent(percent, 1), color = percent),
    hjust = -0.2) +
  theme_hfv() +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  scale_color_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  labs(title = "Percent of developable land dedicated to single-family detached only") +
  theme(axis.text.x = element_blank())

```

```{r}
#| label: fig-bipoc-sfd
#| fig-cap: "Percentage of SFD-only land compared to BIPOC population"
#| eval: false


# Read in data frame 
# hr_bipoc <- read_rds("data/hr/hr_bipoc.rds") |> 
#   st_drop_geometry() |> 
#   arrange(name_long) |> 
#   mutate(bipoc = percent) |> 
#   select(fips, name_long, bipoc)
# 
# hr_sfd <- read_rds("data/hr/hr_vza_nogeo.rds") |> # Read in the data.
#   select(jurisdiction, overlay, sfd, area) |> # Select necessary data fields.
#   filter(overlay == 0) |> # Remove overlay districts.
#   mutate(jurisdiction = case_when(
#     jurisdiction == "Smithfield" ~ "Isle of Wight",
#     jurisdiction == "Windsor" ~ "Isle of Wight", 
#     jurisdiction == "Surry (town)" ~ "Surry",
#     TRUE ~ jurisdiction
#   )) |> 
#   group_by(jurisdiction, sfd) |> # Sum the total acres dedicated to single-family detached in each district.
#   summarise(area = sum(area)) |> 
#   group_by(jurisdiction) |> 
#   mutate(percent = area/sum(area)) |> # Calculate what percentage of total developable area is SFD.
#   filter(sfd == "t") |> 
#   ungroup() |> 
#   add_row(jurisdiction = "York", sfd = "f", area = 0, percent = 0)
# 
# hr_sfd_arrange <- hr_sfd |> 
#   arrange(jurisdiction)
# 
# sfd_bipoc <- cbind(hr_bipoc, hr_sfd_arrange) |> 
#   mutate(sfd = percent) |> 
#   select(jurisdiction, bipoc, sfd)
# 
# # |> 
# #   pivot_longer(4:5,
# #                names_to = "metric",
# #                values_to = "pct") |> 
# #   mutate(metric = case_when(
# #     metric == "bipoc" ~ "BIPOC population (% of total)",
# #     metric == "sfd" ~ "Single-family detached only (% of total developable)"
# #   ))
#   
# 
# ggplot(sfd_bipoc,
#        aes(x = bipoc,
#            y = sfd)) +
#   geom_point() +
#   geom_smooth(method=lm)
# 
# +
#   coord_flip() +
#   theme_hfv() +
#   scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.05))) +
#   theme(legend.position = "top") + 
#   scale_fill_hfv() +
#   labs(title = "Percentage of SFD-only land compared to BIPOC population")



```


### By-Right Housing

```{r}
#| label: tbl-hr-byright
#| tbl-cap: "Percent of developable land that welcomes housing type by-right"

# Create a table that shows the percentage of land that allows for each type of treatment by locality
# https://frontierinstitute.org/reports/the-montana-zoning-atlas-2-0/ use this as an example 

hr_byright_tbl <- read_rds("data/hr/hr_vza_nogeo.rds") |> 
  select(jurisdiction, abbrvname, overlay, 6:9, area) |> 
  filter(overlay == 0) |> 
  group_by(jurisdiction) |> 
  mutate(total_area = sum(area)) |> 
  mutate(family1_pct = case_when(
    family1_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0
  ))|> 
  mutate(family2_pct = case_when(
    family2_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0 
  )) |> 
  mutate(family3_pct = case_when(
    family3_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0 
  )) |> 
  mutate(family4_pct = case_when(
    family4_treatment == "allowed" ~ area/total_area,
    TRUE ~ 0 
  )) |> 
  group_by(jurisdiction) |> 
  summarise(family1_byright = percent(sum(family1_pct)),
            family2_byright = percent(sum(family2_pct)),
            family3_byright = percent(sum(family3_pct)),
            family4_byright = percent(sum(family4_pct))
            ) 


hr_byright_tbl |>
  kable(col.names = c("Jurisdiction", "Single-family", "2-family", "3-family", "4+-family"), 
        align = "lcccc") |> 
  kable_styling(bootstrap_options = c("striped", "condensed"))

```

```{r}
#| label: fig-2plus
#| fig-cap: "Percent of developable land allowing for 2+ family housing by-right" 

# Create a bar graph that shows the total amount of zoned land that welcomes 2+ family housing by-right

hr_byright <- read_rds("data/hr/hr_vza_nogeo.rds") |> 
  select(jurisdiction, abbrvname, overlay, 6:9, area) |> # Select data fields.
  filter(overlay == 0) |> # Filter out overlay districts.
  group_by(jurisdiction) |> 
  mutate(total_area = sum(area)) |> # Calculate the total area of zoned land.
  filter(family2_treatment == "allowed" & family3_treatment == "allowed" & family4_treatment == "allowed") |> 
  group_by(jurisdiction) |> 
  mutate(pct_2plus = area/total_area) |> 
  group_by(jurisdiction) |> 
  summarise(pct_2plus = sum(pct_2plus)) |>
  add_row(jurisdiction = "Smithfield", pct_2plus = 0)
  

ggplot(hr_byright,
       aes(x = reorder(jurisdiction, pct_2plus),
           y = pct_2plus,
           fill = pct_2plus)) +
  geom_col() +
  geom_text(aes(label = percent(pct_2plus, 1), color = pct_2plus),
    hjust = -0.2) +
  coord_flip() +
  theme_hfv() +
  scale_fill_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  scale_color_gradientn(colours = c("#ADDCB1", "#46A88F", "#207A88", "#0F1A4E")) + # Use custom HFV gradient.
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Percent of developable land allowing for 2+ family housing by-right")  +
  theme(axis.text.x = element_blank())


```




### Minimum lot sizes

```{r}

# What is the average minimum lot size for each type of treatment? 
# Blank fields should be treated as NA and should not be included in the mean calculation.


hr_lots <- read_rds("data/hr/hr_vza_nogeo.rds")|> 
  select(jurisdiction, abbrvname, overlay, hr_custom_fields_min1, hr_custom_fields_min2,
         hr_custom_fields_min3, hr_custom_fields_min4) 


```


## Accessory Dwelling Units

```{r}
#| label: fig-hr-adus
#| fig-cap: "Percent of developable land that allows for ADUs by-right"

# Create a table that shows the percentage of land that allows for ADUs by-right
# https://frontierinstitute.org/reports/the-montana-zoning-atlas-2-0/ use this as an example 

hr_adu <- read_rds("data/hr/hr_vza_nogeo.rds") 

```


Many localities across Hampton Roads address ADUs in their zoning ordinances, while a few (Chesapeake and Newport News) make no mention of them at all. 
```{r}
#| label: fig-hr-adu-restrictions
#| fig-cap: "Summary of ADU regulations in Hampton Roads"

# Create a table for the following csv.
# https://frontierinstitute.org/reports/the-montana-zoning-atlas-2-0/ use this as an example 

hr_adu_summary <- read_csv("data/hr/hr_adu_summary.csv")

hr_adu_summary |> 
  kable() |> 
  kable_styling(bootstrap_options = c("striped", "condensed"))



```






